\documentclass[notitlepage]{report}
\usepackage{graphicx}
\usepackage[parfill]{parskip}
\usepackage{hyperref}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{amsmath}
\usepackage{listings}


\usepackage{etoolbox}
\makeatletter
\patchcmd{\chapter}{\if@openright\cleardoublepage\else\clearpage\fi}{}{}{}

\title{CS444/644 Assignment 1 Technical Report}
\author{Ho-Yi Fung, Jianchu Li, Zhiyuan Lin}
\date{\today}

\begin{document}

{\let\newpage\relax\maketitle}
\begin{abstract}
The purpose of this assignment is to design and implement the lexical and syntactic analysis module of a Joos 1W compiler. The original objective has been successfully achieved. In this report we discuss in detail the design and implementation of our program, and the issues that the team faced in the process.
\end{abstract}


\tableofcontents



%\begingroup
%\let\clearpage\relax
\include{intro}

\include{design}

\include{implementation}

\include{testing}
%\endgroup

\chapter{Conclusion}
The implementation of lexical and syntactic analysis for our compiler has been a success. All the modules were implemented as designed and passes all the test cases. There are however improvements we could make to the program. Time complexity was always carefully considered when designing the program, but space efficiency could be improved. This was a conscious choice made by the development team as it was believed that any modern machine would have enough memory for these operations. The AST nodes have been implemented but further development and modification are expected to make it easier to use for later phases. Any future changes for this module will be documented in the next report.

\bibliographystyle{ieeetr}
\bibliography{a1}

\end{document}